# Cartography for Epidemiology I

## w2 Learning objectives

```{r echo=FALSE, message=F, error = F, results = 'hide'}
library(huxtable)
library(tidyverse)
lo <- rio::import('../schedule.xlsx', which = 'learning objectives')
```

```{r echo = F}
## define section
x <- 'Cartography for Epidemiology I'
lo %>%
  filter(Mod2 == x) %>%
  as_hux() %>%
  select(3) %>%
  set_all_padding(2) %>%
  set_label('learning-ob') %>%
  set_caption('Learning objectives by weekly module') %>%
  set_bold(1, everywhere) %>%
  set_width(1) %>%
  set_wrap(TRUE) %>%
  theme_orange() 
```

## w2 Additional Resources 

## w2 Spatial Thinking in Epidemiology

* Matching epidemiologic needs and objectives to specific attribute and geography data
* Matching cartographic representation to the *meaning* of data (L2d)

## w2 Spatial Analysis in Epidemiology 

Goal in this section is focus on data preparation and geo-processing for epidemiology

* `tigris` and census geography
* `tidycensus` to acquire ACS/census data +/- geography
* Review tips for `dplyr` including mutate, select, groupby, summarize


