# (PART) Appendices {-}


# Tips for reproducibility in `R`

## Additional Resources {-}

* [R Markdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [Comprehensive guide to using R Markdown](https://bookdown.org/yihui/rmarkdown/) 
  + [Chapter within the R Markdown guide specific to Notebooks](https://bookdown.org/yihui/rmarkdown/notebook.html)
* [Working with *Projects* in `R`](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)
* [R for Data Science - Workflow Projects](https://r4ds.had.co.nz/workflow-projects.html)


## Workflows to enhance reproducibility {-}

Because `R` and RStudio are often used for data preparation, analysis, and reporting, the fundamental importance of *reproducibility* (making analytic processes transparent, interpretable and repeatable) is built-in through many features.  This Appendix introduces several strategies that are important for reproducibility broadly, and also important for the work you do in this course. First, there is a brief introduction to *projects* in RStudio, and then there is a slightly more in-depth description of a specific file format, `rmarkdown` and how it can be used to create Notebooks.

### Using Projects in `R` {-}

A **project** in `R` functions in terms of organization the same way you may use *folders* on your computer to sort and separate into some logical scheme. In other words, it is a place where you put multiple documents or files that are related to one another. For instance, you might choose to have a single project for each week of this class, and perhaps a separate project for each assignment. In each *project directory* (folder) you could store the data, the scripts or code, and any outputs (e.g. saved maps or other saved objects) that are specific to that week or assignment.

The advantage of creating a formal *project* in RStudio (rather than just a regular folder, for example), is that RStudio projects have certain benefits for your coding workflow. 

* When you open a project, the *working directory* (e.g. the root directory or file path where `R` looks for files when you import) is automatically set to be inside the project folder. This means that if you keep your data inside the project, you will never have to worry about broken links or incorrect file paths that occur because data was moved.
* Projects remember environmental settings in RStudio, so you may customize something to a specific project and that will be remembered each time you open the project.
* If you ever work with a version control system such as Github, projects are the natural strategy to contain a repository

**To create a new project**:

1.Look in the upper-right corner of RStudio for the blue-ish R symbol that will likely say '*Project*'. Click the pull-down menu and select New Project
2. You will see the Project Wizard open with three options:
  + If you have not yet created the folder on your computer that will be your project, choose *New Directory*
  + If you already have a folder (e.g. perhaps it is named 'Week1'), choose *Existing Directory*
  + If you are are forking or checking out a repository from Github, GitLab or other system, choose *Version Control*
3. Navigate to the location you want your *new folder* to be, or else the location where you *existing folder* already is
4. Name the project and click Create Project


Once the project is created, you can navigate via your finder to that folder. You will notice a new file with extension `.Rproj`. If you double-click this file, your project will open, including whatever files and settings you have already worked on.

## Understanding directory and file path names in `R` {-}

Recall that in `R` there is always a defined *Working Directory*, which is the *home* for `R` in this current working session. If you are using *Projects* as defined above (*which I highly recommend*), then the working directory will always be the main folder you defined when you created *this project*.  

If you are not using a *Project*, and you wondered where your home directory is you could do this:

```{r}
getwd()
```

While in general I **do not recommend** doing this, it is also possible to change your working directory by setting it to a new path like this:

```{r eval = F}
setwd('H:/mkram02/gis-files') # this changes my working directory to be on the H:/ drive
```

```{block2, type = 'rmdcaution'}
Note the direction of the backslash in the path above! If you work in a Windows OS environment, then you should be careful if you copy/paste a pathname, because the direction of the backslash within pathanmes in *Windows* is the __opposite__ of what is used in `R`! That is because `R` is more aligned with Unix styles, and Windows is not. If you work within a Mac OS environment then you are all set! Mac is also Unix-aligned. If you are in Windows, you must either manually reverse the backslashes after you copy/paste, or else you can use double backslashes like this: `H:\\mkram02\\gis-files`.
```

In general it is easiest if the data you want to read or import (e.g. spatial or aspatial datasets), and the files you want to save or export (e.g. results, figures or maps, intermediate datasets) are simply in the working directory. That way you never need a path name; you only need the file name.

However, inevitably you will need to load or save something that is not in the root of your working directory. For example it often makes sense to have a separate folder within your *Project* directory to store all plots, maps, and images (e.g. I often name this something obvious like `images` or `plots`). Similarly, you may have multiple datasets and it is easier to put all of them into a sub-folder labeled `data`.  When your target file (either for loading or saving) is *not in the working directory*, you must specify *where it is*.

### Specify a location within the working directory {-}

If you have sub-folders within your working directory (e.g. within your Project folder for example), it is straightforward to locate them simply by typing the path to find them. 

```{r eval = F}
# This loads a file called 'georgia.csv' contained within my data folder, and
# specifically within a sub-folder to data called 'death-data'
dd <- read.csv('data/death-data/georgia.csv')
```

```{block2, type = 'rmdtip'}
A handy trick for navigating nested folders when you are referencing a path name is to use the `tab` key as you type the path. For instance if you have an `R` function that requires a file location, just type the opening quote, and hit the `tab` key on your computer. `R` will auto-populate all of the files and folder and you can click on the one you want. Once you are __in__ that folder, hit `tab` again and it will auto-populate the files/folders in that folder. You can make it go faster by beginning to type the first few letters.
```

### Specify a relative location outside the working directory {-}

What if you have one folder for this entire course, and inside it you have a separate project directory for each week. If you are working on the project for `Week2`, you might wish to load a file that you saved previously in `Week1`. In other words it is not a sub-folder, but is actually *outside* of the current directory. You could use the `setwd()` function to change the location, but that creates a possibly fragile *absolute pathname* and can be dangerous. Instead you could create a more robust *relative pathname* by referring to the other file in relation to your current location.

Using two dots in a pathname tells `R` to **go up a level** in the directory. So to if the `georgia.csv` file I referred to above were in your `Week1` directory, but your are currently woring in `Week2` you could do this:

```{r eval = F}
dd <- read.csv('../data/death-data/georgia.csv')
```

This means "*go up a level, then look in the data folder, then the death-data folder, then load the georgia.csv file*".  If you need to go up two (or more) levels, simply repeat:  `../../data/death-data.georgia.csv`

## Why `R Notebooks`? {-}

For most assignments in this course, at least a portion of the deliverable will be a fully-functional, annotated `R Notebook`.  These *notebooks* are actually a specific case of `rmarkdown` which itself is a format for creating reproducible documents with interspersed `R` code, analytic results and text. For example this eBook, and many other resources in this course are created using `rmarkdown` or related packages such as `bookdown`. 

But as I said, `R Notebooks` are a specific instance or case of markdown that is incorporated into R Studio and has some nice features for the applied data analyst.

* Notebooks contain text that explains what is happening, interprets findings, or notes areas in need of further exploration
* Notebooks contain functional `R` code that when run places the results **inside** the Notebook document
* Notebooks work in an *interactive* mode. This means that as you are coding and working you can see the results in the document. When you save the Notebook the text, the code **and the results** are saved! 

So the reason for using Notebooks is that they provide a means for clear annotation and documentation combined with ready reproducitiblity. Reproducibility means that someone else (or a future you!) could come back and get the same result again.

To benefit from the advantages above, I recommend you gain familiarity with the basic (and perhaps the optional) formatting described below. I also recommend you develop a knack for rich annotation and documentation, not just brief (often cryptic) comments that we are all used to writing in SAS and other code! Document what you **plan to do**. Document what you **did**.  Document what the **results means**. Document what else **needs to be done**.

## What you need to know {-}

This file summarizes both **important** and just a small handful of **optional** functions for effectively using `R` Notebooks. The **important** functions are those necessary to effectively intersperse narrative text and description communicating what you did and what it means, with clear R code and the resulting output.  The **optional** parts are about some simple formatting tools. These **are not** necessary for your homework (our goal is documentation of analytic process not being 'pretty'), but you may find them useful.

## Important R Notebook functions {-}

### The YAML {-}

```
---
title: "Title of your notebook"
author: "Your Name Here"
date: "Submission date here"
output:
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
---
```


When you create a new `R` Notebook or `R` Markdown file from within R Studio (e.g. via *File > New File > R Notebook*), a 'YAML' will automatically be created at the top of the script delineated by three dash lines `---`.  *YAML* stands for "*yet another markup language*" and it is a set of instructions about how the finished notebook will look and be structured. You can accept the default YAML structure (of course modifying the title) or copy/paste the YAML from the top of this script. You can also read more online about additional customizations to the YAML, but none are necessary for this course.

## Typing text {-}

The utility of `R` Notebooks is the ability to more completely document your thinking and your process as you carry out analyses. It is not necessary to be wordy just for the sake of taking up space, but this is an opportunity to clearly delineate goals, steps, data sources, interpretations, etc.  

You can just start typing text in the script to serve this purpose. Some text formatting functions are summarized later in this document, and in Cheatsheets and online resources linked to elsehwhere.



## Adding `R` Code {-}

`R` Notebooks let you write `R` code within your Markdown file, and then run that code, seeing the results appear right under the code (rather than only in the Console, where they usually appear). 

There are 2 ways to add a new chunk of R code:

1. Click the green `C-Insert` button at the top of the editor panel in `R` Studio. The top option is for `R` code.  
2. Use a keyboard short cut: 
    + *Mac*  Command + Shift + I
    + *Windows* Ctrl + Alt + I
    
Notice these `R` code chunks are delineated by three back-ticks (sort of like apostrophers)...these back-ticks are typically on the same key as the tilde (~) on the upper left of most keyboards. The space between the sets of 3 back-ticks is where the `R` code goes and this is called a *code chunk*.

Inside these 'chunks' you can type `R` code just as you would in a regular `R` script.

When you want to run it you can either:

1. Place your cursor on a line and click Ctrl+enter (Windows) or CMD+Return (Mac), or you can click the *Run* button at th etop of the editor pane in R Studio. 
2. To run all of the code within a chunk click the green *Run Current Chunk* button at the upper-right of the code chunk. 

Below is some code and the corresponding results.


```{r}
head(mtcars)
```

```{r}
plot(cars)
```


In this way you can iterate through your analytic process...switching between running code, viewing output, documenting in free text.

If you want to see what your current work-in-progress looks like in HTML, you can click the *Preview* button at the top of the panel. This will both save your document, and open the *Viewer* panel.


## Making tables {-}

While not required, you may want to summarize data in a table in `R` Markdown. There are packages devoted to creating tables, but you can create a quick-and-dirty table just using keyboard symbols.  

1. First start by making a header row. Separate each column name with a *'pipe'* symbol, `|`
2. Put a continuous line of dashes (`-----`) under each column name, separating columns with pipe symbole (`|`)
3. Now type text corresponding to each row and column. Separate columns with pipe (`|`) and separate rows by carriage return/Enter

So the following text typed directly into the Markdown file (e.g. not inside a code chunk):

```
Column 1  | Column 2 | Column 3
----------|----------|-----------
Text 1    | Text 2   | Text 3
Next line | Next line 2 | Next line 3
```

Will produce the following output:

Column 1  | Column 2 | Column 3
----------|----------|-----------
Text 1    | Text 2   | Text 3
Next line | Next line 2 | Next line 3

## Workflow {-}

The benefit of Notebooks (slightly different from regular Markdown) is that you can work interactively with your code, seeing results immediately just as you would with a regular script. In contrast a 'regular' Markdown file doesn't run the code until you click 'Knit'.

Here is what I recommend for workflow:

1. Name your script and save it to the desired location. Wherever you save it will become the default Working Directory for any code run from *within the Notebook*.  
2. Carry out your analysis, inserting code chunks, running them, and documenting them as you go.  
3.  As a final check of *reproducibility* (the assurance that your code is self-contained and not dependent on steps you did outside the script) I recommend you always end by clicking the *RUN* button at the top of the panel. Specifically, choose **Restart R and Run all Chunks**.  If there is an error when you did this, then something is missing in your code.

## Optional functions {-}

The list of formatting functions is long. I include only a couple I find useful (but not mandatory) here:

## Customizing your YAML {-}

While the default YAML is perfectly fine, the YAML at the top of this script includes a few added functions including:  

1. Specify a table of contents - this only works if you use headers
2. Specify section numbering
3. Specify that the table of contents should be 'floating' which means that in *html* it is visible even when you scroll. For PDF rendering, 'float' is not an option. 


## Simple formatting of your Notebook {-}

Often it is helpful use headers to separate tasks or steps in your code. You can easily create headers using the hastag/pound sign `#`. Specifically...

* `#` at the beginning of the line denotes a top-level (level-1) header that will be large and bold.
* `##` at the beginning of the line denotes level-2 header
* `###` unsurprisingly is a level-3 header!
* Make sure there is a space between the `#` and the text
* Always leave a blank line (return/enter) between the header text and the '*regular*' text.

You can also make numbered or bulleted lists if that is helpful.  A line that begins with either an asterisk (`*`) or a number will begin a bulleted or numbered list.

Headers are populated into the table of contents, if specified.

## Text formatting {-}

The [`R` Markdown Cheatsheets](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) have lots of examples of formatting. Three things that I use more frequently are bold, italics, and numbered or bulleted lists.  

Key stroke            | Result
------------          | -------
```*italics*```       | *italics*
```**bold**```        | **bold**


1. Numbered lists start with number, and each line must end with 2 space (or have blank line between).
2. Instead of numbers you can use letters

* Bulleted lists can be initiated with an asterisk or +, and also must have 2 spaces (or blank carriage return) at end of each item.


## Final Note {-}

Remember that a final step when you think you are done with a project, is to Click `Restart R and Run all Chunks`, and then save/preview the Notebook **after** doing this to be sure it is what you expect. 



