# Tips for working with `sf` data class {#sf-overview}

## `st_set_geom()`

There is a feature of `sf` class data in that the special column containing the *geometry* information (often labeled `geom`) is different from other variables. Specifically it is **sticky**. *Stickiness* in a variable means that as you manipulate an `sf` data object, the `geom` column almost always *sticks* to the rest of the data even when you try to remove it.

Imagine what would happen in a regular `data.frame` if you typed this code into the console `mvc[1, 1:2]`.  Typically that kind of numerical indexing would cause `R` to return *row 1* for *columns 1 to 2*. However, when we try this in `R` with an `sf` object this is what happens:

```{r}
library(sf)
library(tidyverse)

mvc <- st_read('../DATA/GA_MVC/ga_mvc.gpkg')

mvc[1, 1:2]
```


Notice that we did get the first row, and the first and second column but we *also got the `geom` column* even though we didn't request it.  This *stickiness* is generally desirable, because it is so important to keep geographic/geometry data connected to attribute data. However there are times when we want to drop that information. There are several ways to do so, but here is the most explicit way:

```{r}
mvc2 <- st_set_geometry(mvc, NULL)
```

This literally erases or sets to `NULL` the geometry column. It cannot be retrieved without going back to the original data.

```{r}
# look at the class of the original and the modified object
class(mvc)
class(mvc2)

# look at the first row and 1-2nd column after NULLing geom
mvc2[1, 1:2]
```


## `st_as_sf()`

There are also times when, inextricably, your data set that **seems** like an `sf` object gets rejected by a function as not having geometry information or not being `sf`.  Sometimes data manipulation steps strip away the `sf` data class even though the `geom` column still exists. When this happens you can *reinstate* the class status by calling `st_as_sf()`. Essentially this is a formal way for declaring an object to be `sf` by explicitly defining the *spatial* component.

```{r}

```


